Expert-In-The-Loop Causal Discovery: Iterative Model Refinement Using Expert Knowledge

### 2025-02-08 paper/main.tex
Updates

\subsection{Integrating domain knowledge in causal discovery}

@@

We therefore consider the following two models of a domain experts: Given two variables
\emph{X} and \emph{Y} that are part of an unknown causal DAG $G$, an \emph{strong ancestral oracle} 
$\mathcal{A}_G$ is defined as:
$$\mathcal{A}_G(X,Y)=\begin{cases}
 X \to Y & \textrm{if } X \in \textrm{An}_G(Y) \\
 X \gets Y & \textrm{if } Y \in \textrm{An}_G(X) \\
 \textrm{None} & \textrm{otherwise} \\
\end{cases}$$

@@

\begin{figure}[t!]
	\begin{subfigure}{0.125 \textwidth}
		\centering
		\includegraphics[page=1]{example.pdf}
		\caption{True DAG}
	\end{subfigure}%
	\begin{subfigure}{0.125 \textwidth}
		\centering
		\includegraphics[page=2]{example.pdf}
		\caption{No edges.}
	\end{subfigure}%
	\begin{subfigure}{0.125 \textwidth}
		\centering
		\includegraphics[page=3]{example.pdf}
		\caption{$ X_1 \rightarrow X_4 $}
	\end{subfigure}%
	\begin{subfigure}{0.125 \textwidth}
		\centering
		\includegraphics[page=4]{example.pdf}
		\caption{$ X_1 \rightarrow X_2 $}
	\end{subfigure}
	\begin{subfigure}{0.125 \textwidth}
		\centering
		\includegraphics[page=5]{example.pdf}
		\caption{$ X_1 \rightarrow X_3 $}
	\end{subfigure}%
	\begin{subfigure}{0.125 \textwidth}
		\centering
		\includegraphics[page=6]{example.pdf}
		\caption{$ X_2 \rightarrow X_4 $}
	\end{subfigure}%
	\begin{subfigure}{0.250 \textwidth}
		\centering
		\includegraphics[page=7]{example.pdf}
		\caption{$ X_3 \rightarrow X_4 $. $ X_1 \not \rightarrow X_4 $.}
	\end{subfigure}
	\caption{An example showing each iteration of the \emph{ExpertInLoop} algorithm. The dashed edges show all the potential edges that \emph{Expand} step could add in each iteration.}
\end{figure}

@@

In this section, we give some of the practical aspects of the
\emph{ExpertInLoop} algorithm, and then compare the algorithm with automated
causal discovery algorithms.

@@

\subsection{Comparison to Automated Algorithms}
\begin{figure}[t!]
	\centering
	\begin{subfigure}{0.5\textwidth}
		\centering
		\includegraphics{../code/fig3/shd_ribbon.pdf}
		\caption{}
	\end{subfigure}
	\begin{subfigure}{0.5\textwidth}
		\centering
		\includegraphics{../code/fig3/sid_ribbon.pdf}
		\caption{}
	\end{subfigure}
	\caption{Comparison of PC, Hill-Climb Search, and GES algorithms against
		\emph{ExpertInLoop} algorithm. As automated algorithms only
		recover the CPDAG, we use the best and worst scoring
		orientation of the CPDAG to get the range. We test
		\emph{ExpertInLoop} with varying values of expert accuracy, $ \alpha = \{0.1, 0.3, 0.5, 0.7, 0.9\} $. The corresponding
		$\alpha_{\textrm{eff}} $ is shown in the plot.}
	\label{fig:shd_sid}
\end{figure}

@@

The \emph{ExpertInLoop} algorithm requires two oracles, one that can answer
d-separation queries and an ancestral oracle that can give the orientation or
non-existence of edges between pairs of variables. For answering the
d-separation queries, we use a CI test, the partial correlation test with a
p-value threshold of $ 0.05 $. Secondly, to tell the direction between edges or
non-existence of edges, we simulate an expert with accuracy $ \alpha $ as
follows:

@@

We analyzed the performance of \emph{ExpertInLoop} algorithm by comparing it
with three other automated algorithms: PC, Hill-Climb Search, and Greedy
Equivalence Search (GES). For the analysis, we simulated $ 500 $ samples from
known DAGs and use the algorithms to recover the original DAG. We start by
generating a random DAG on $10$ nodes and use linear models with random effects
to generate the data. We vary the edge probability of the DAG to generate DAGs
with varying density. We repeat the data generation $30$ times for each density
value for the DAG. We compare the learned DAG to the original DAG using two
metrics: Structural Hamming Distance (SHD) and Structural Intervention Distance
(SID) \citep{Peters2015}.

@@

As the automated algorithms can only recover the CPDAG from data, and our
method recovers the DAG. To compare them fairly, we computed the SHD and SID
for all possible orientations of the CPDAG to show the best and worst case
scenarios. The results are shown in Figure~\ref{fig:shd_sid}. For SHD, the
performance of \emph{ExpertInLoop} is comparable to the automated algorithms
for $ \alpha = 0.3 (\alpha_{\textrm{eff}} = 0.53 $, and performs better for
higher $ \alpha $ values. For SID, the \emph{ExpertInLoop} performs better in
denser DAGs.

>>>>

\subsection{Integrating domain knowledge in causal discovery}
Multiple ways to integrate, originally using forbidden and required edges. Or
temporal algorithms by specifying causal tiers.

@@

We therefore consider the following two models of a domain experts: Given two variables
$ X $ and $ Y $ that are part of an unknown causal DAG $G$, an \emph{strong ancestral oracle} 
$\mathcal{A}_G$ is defined as:
$$\mathcal{A}_G(X,Y)=\begin{cases}
 X \to Y & \textrm{if } X \in \textrm{An}_G(Y) \\
 X \gets Y & \textrm{if } Y \in \textrm{An}_G(X) \\
 \textrm{None} & \textrm{otherwise} \\
\end{cases}$$

@@

\begin{figure}[t!]
	\begin{subfigure}{0.125 \textwidth}
		\centering
		\includegraphics[page=1]{example.pdf}
		\caption{True DAG}
	\end{subfigure}%
	\begin{subfigure}{0.125 \textwidth}
		\centering
		\includegraphics[page=2]{example.pdf}
		\caption{No edges.}
	\end{subfigure}%
	\begin{subfigure}{0.125 \textwidth}
		\centering
		\includegraphics[page=3]{example.pdf}
		\caption{$ X_1 \rightarrow X_4 $}
	\end{subfigure}%
	\begin{subfigure}{0.125 \textwidth}
		\centering
		\includegraphics[page=4]{example.pdf}
		\caption{$ X_1 \rightarrow X_2 $}
	\end{subfigure}
	\begin{subfigure}{0.125 \textwidth}
		\centering
		\includegraphics[page=5]{example.pdf}
		\caption{$ X_1 \rightarrow X_3 $}
	\end{subfigure}%
	\begin{subfigure}{0.125 \textwidth}
		\centering
		\includegraphics[page=6]{example.pdf}
		\caption{$ X_2 \rightarrow X_4 $}
	\end{subfigure}%
	\begin{subfigure}{0.250 \textwidth}
		\centering
		\includegraphics[page=7]{example.pdf}
		\caption{$ X_3 \rightarrow X_4 $. $ X_1 \not \rightarrow X_4 $.}
	\end{subfigure}
	\caption{An example showing each iteration of the \textsc{ExpertInLoop} algorithm. The dashed edges show all the potential edges that \textsc{Expand} step could add in each iteration.}
\end{figure}

@@

In this section, we give some of the practical aspects of the
\textsc{ExpertInLoop} algorithm, and then compare the algorithm with automated
causal discovery algorithms.

@@

\subsection{Comparison to Automated Algorithms}
\begin{figure}[t!]
	\centering
	\begin{subfigure}{0.5\textwidth}
		\centering
		\includegraphics{../code/fig3/shd_ribbon.pdf}
		\caption{}
	\end{subfigure}
	\begin{subfigure}{0.5\textwidth}
		\centering
		\includegraphics{../code/fig3/sid_ribbon.pdf}
		\caption{}
	\end{subfigure}
	\caption{Comparison of PC, Hill-Climb Search, and GES algorithms against
		\textsc{ExpertInLoop} algorithm. As automated algorithms only
		recover the CPDAG, we use the best and worst scoring
		orientation of the CPDAG to get the range. We test
		\textsc{ExpertInLoop} with varying values of expert accuracy, $ \alpha = \{0.1, 0.3, 0.5, 0.7, 0.9\} $. The corresponding
		$\alpha_{\textrm{eff}} $ is shown in the plot.}
	\label{fig:shd_sid}
\end{figure}

@@

The \textsc{ExpertInLoop} algorithm requires two oracles, one that can answer
d-separation queries and an ancestral oracle that can give the orientation or
non-existence of edges between pairs of variables. For answering the
d-separation queries, we use a CI test, the partial correlation test with a
p-value threshold of $ 0.05 $. Secondly, to tell the direction between edges or
non-existence of edges, we simulate an expert with accuracy $ \alpha $ as
follows:

@@

We analyzed the performance of \textsc{ExpertInLoop} algorithm by comparing it
with three other automated algorithms: PC, Hill-Climb Search, and Greedy
Equivalence Search (GES). For the analysis, we simulated $ 500 $ samples from
known DAGs and use the algorithms to recover the original DAG. We start by
generating a random DAG on $10$ nodes and use linear models with random effects
to generate the data. We vary the edge probability of the DAG to generate DAGs
with varying density. We repeat the data generation $30$ times for each density
value for the DAG. We compare the learned DAG to the original DAG using two
metrics: Structural Hamming Distance (SHD) and Structural Intervention Distance
(SID) \citep{Peters2015}.

@@

As the automated algorithms can only recover the CPDAG from data, and our
method recovers the DAG. To compare them fairly, we computed the SHD and SID
for all possible orientations of the CPDAG to show the best and worst case
scenarios. The results are shown in Figure~\ref{fig:shd_sid}. For SHD, the
performance of \textsc{ExpertInLoop} is comparable to the automated algorithms
for $ \alpha = 0.3 (\alpha_{\textrm{eff}} = 0.53 $, and performs better for
higher $ \alpha $ values. For SID, the \textsc{ExpertInLoop} performs better in
denser DAGs.
